---
title: "My Life"
author: "DK"
date: "2025-06-02"
categories: [life]
---

Attended a gathering.

The gathering organized by the `Empty Nesters`

The mean and the median age of attendees was 70 years of age.

The organizer and the host of the evening introduced the speaker,

> Often we learn about the unknown part of a person on their funeral...

The speaker said,

> I am glad this is not my funeral today
>
> My life has been different than most
>
> I hope you learn something from it.

The speaker and his wife celebrated their 60th wedding anniversary last year.

We first met them, when we moved into the area some 25 years ago.

```{r}
#| echo: false
#| message: false
#| warning: false
library(leaflet)

# Create leaflet map
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -98.76019924518442, lat = 20.100969616918896, popup = "Pachuca Hidalgo")
```

His life started in the Hidalgo State in Mexico.

3 women influenced his life.
His aunt, his mom and his wife

> There is nothing better on the face of the earth, the gospel



## Museum of Mormon Mexican History

He and his wife found a calling in life.
A museum dedicated to the Mormon Saints in Mexico.


```{r}
#| echo: false
#| message: false
#| warning: false
library(leaflet)

# Create leaflet map
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -111.65627966851787, lat = 40.25390357966943, popup = "Museum of Mormon Mexican History")
```
<https://museumofmormonmexicanhistory.com/about-us>
